A Bayesian search to find \\high-mass black holes in LIGO data

### 2020-08-19 main.tex
work on methods

\newcommand{\prior}{{\sc prior}\xspace}
\newcommand{\bilby}{{\sc Bilby}\xspace}
\newcommand{\bilbypipe}{{\sc bilby\_pipe}\xspace}
\newcommand{\bilbypipegdb}{\texttt{bilby\_pipe\_gracedb}\xspace}
\newcommand{\pbilby}{{\sc pBilby}\xspace}
\newcommand{\dynesty}{{\sc dynesty}\xspace}
\newcommand{\cpnest}{{\sc cpnest}\xspace}
\newcommand{\ptemcee}{{\sc ptemcee}\xspace}
\newcommand{\gwpy}{{\sc GWpy}\xspace}
\newcommand{\imrphenomp}{{\sc IMRPhenomPv2}\xspace}
\newcommand{\pycbc}{{\sc PyCBC}\xspace}
\newcommand{\msun}{{M${}_\odot$}\xspace}

\begin{document}

@@

One method to account for glitches while searching for gravitational waves from coalescing compact binaries is by utilising a signal consistency test \cite{pycbc_short_duration_transients} or a signal coherence test amongst a network of detectors \cite{bci, bcr_paper,  bcr_gw151216, bayesian_odds}. A signal-coherence test such as the Bayesian coherence ratio (BCR) \cite{bcr_paper} provides an odds ratio comparing the probability that the data contains an astrophysical signal versus the probability that the data contains an incoherent glitch or noise. Although an odds ratio, the BCR can be used  as a ranking statistic to calculate the statistical significance of candidate gravitational-wave signals from compact binary coalescences in the detector data \cite{bcr_paper}.

In this paper we utilise the BCR to calculate $p_\text{astro}$, the probability that a signal is of astrophysical origin [cite],  for all the high-mass (systems with total masses in the range of $55-500$ \msun) candidate gravitational-wave signals from compact binary coalescences in the detector data recorded during aLIGO's second observing run (O2).  We find that (a) our search does not identify any new stellar mass or intermediate mass black holes; (b) high-mass events reported in the GWTC-1, including GW170729 (an event with disputed $p_\text{astro}$ amongst various search pipelines) have high $p_\text{astro}$; and that (c) high-mass events detected from independent groups have low $p_\text{astro}$.

\section{PyCBC}
% PyCBC asks for the 5 citations:
% http://pycbc.org/pycbc/latest/html/credit.html#searches-for-compact-binary-coalescence

 such as \pycbc \cite{pycbc_code} have tried to account for these short-duration glitches \cite{pycbc_short_duration_transients}. 
Search algorithms for gravitational waves from coalescing compact binaries, such as PyCBC \cite{pycbc_code}, have tried to account for these short-duration glitches \cite{pycbc_short_duration_transients}. At its core, PyCBC performs a matched-filter search for binary merger signals using a bank of gravitational-wave template waveforms \cite{pycbc_og0, pycbc_og1, pycbc_og2, pycbc_og3, pycbc_og4, pycbc_og5}.  To account for the short-duration glitches,

 to As demonstrated in \citet{blip_glitches},  High-mass mergers have
very few in-band wave cycles, and hence can easily be mistaken for short-duration instrumental transients.

A foreground trigger is a section of LIGO data identified by data searching pipelines to potentially harbour some
gravitational wave signal (more on triggers in Appendix\textasciitilde\ref{searches}). In the ground-based gravitational wave
detectors, many spurious instrumental noise events fool the pipelines into considering the noise events as triggers
(false-positive triggers). These instrumental-triggers are called glitches. Luckily, unlike triggers from astrophysical
sources, glitches do not often occur at the same time or with the same morphology across the various gravitational wave
detectors -- they are often \emph{incoherent} amongst detectors. Additionally, these short-duration glitches appear very
similar to high-mass gravitational-wave signals, making the two challenging to distinguish.

One explanation for the absence of detections may be due to the low merger frequency of the high-mass systems. The ground-based detectors are much less sensitive at lower frequencies due to various noise sources (e.g. thermal, suspension noise). Another difficulty with the low frequencies is the presence of transient short-duration instrumental non-Gaussian features, known as glitches, in the interferometer's recorded data. Some of the short duration glitches can occur as often as a few times every hour~\cite{blipGlitches}. These short-duration glitches are challenging to distinguish from the short-duration gravitational-wave signals that we expect to see from high-mass binary black holes.

The high occurrence rate of short duration glitches down-weight the significance of short duration LIGO foreground triggers, GPS times marking sections of LIGO data to potentially harbour gravitational wave signals.

This paper discusses a novel method to identify and distinguish these high-mass gravitational wave signals from short-duration glitches in gravitational-wave data. It then employs the method on the PyCBC high-mass triggers from O2 and provides frequentistic estimates of the significance of various gravitational-wave candidates and events.

This paper is organized as follows: Section 2 provides an overview of the search
pipeline and the methods used to detect gravitational waves from compact-object
binaries in LIGO data. Section 3 gives a description of the developments implemented
in this pipeline. Section 4 compares the performance of the new pipeline to that of the
pipeline that analyzed the sixth LIGO science run and Virgo’s second and third science
runs [27]. Finally, Section 5 summarizes our findings and suggests directions for future
improvements.

@@

\begin{itemize}
    \item briefly explain PyCBC search
    \item define triggers
    \item Extract PyCBC foreground, background and software injection trigger's GPS times based on template duration of triggers
    \item analyse the triggers using Bayesian Parameter Estimation to determine $\log{Z_s}$, $\log{Z_g}$ and $\log{Z_n}$
    \item calculate BCR
    \item Remove outlier BCRs using median method
    \item Calculate PDFs
    \item Find $\alpha$ and $\beta$ values such that separation of background and software injection PDFs is maximally apart.
    \item use tuned $\alpha$ and $\beta$ from previous step to calculate BCRs for foreground
    \item calculate the $p_{astro}$ of the foreground events to determine the probability of the trigger being from an astrophysical source. 
\end{itemize}

>>>>

\newcommand{\prior}{{\sc prior}\xspace}
\newcommand{\bilby}{{\sc Bilby}\xspace}
\newcommand{\bilbypipe}{{\sc bilby\_pipe}\xspace}
\newcommand{\bilbypipegdb}{\texttt{bilby\_pipe\_gracedb}\xspace}
\newcommand{\pbilby}{{\sc pBilby}\xspace}
\newcommand{\dynesty}{{\sc dynesty}\xspace}
\newcommand{\cpnest}{{\sc cpnest}\xspace}
\newcommand{\ptemcee}{{\sc ptemcee}\xspace}
\newcommand{\gwpy}{{\sc GWpy}\xspace}
\newcommand{\imrphenomp}{{\sc IMRPhenomPv2}\xspace}
\newcommand{\pycbc}{{\sc PyCBC}\xspace}
\newcommand{\bcr}{{\sc BCR}\xspace}
\newcommand{\pastro}{{$p_\text{astro}$}\xspace}
\newcommand{\msun}{{M${}_\odot$}\xspace}
\begin{document}

@@

One method to account for glitches while searching for gravitational waves from coalescing compact binaries is by utilising a signal consistency test \cite{pycbc_short_duration_transients} or a signal coherence test amongst a network of detectors \cite{bci, bcr_paper,  bcr_gw151216, bayesian_odds}. A signal-coherence test such as the Bayesian coherence ratio (\bcr) \cite{bcr_paper} provides a Bayesian odds ratio comparing the probability that the data contains an astrophysical signal versus the probability that the data contains an incoherent glitch or noise. Although an odds ratio, the \bcr can be used  as a frequentistic \textit{ranking statistic}, ranking the likelihood that a trigger is due to a gravitational-wave signal \cite{bcr_paper}. The \bcr as a ranking statistic is much more powerful than a matched-filter SNR as the \bcr incorporates numerous features that may distinguish gravitational waves from noise and demands that signals must be coherent while glitches must be incoherent amongst different detectors.

In this paper we utilise the \bcr to calculate \pastro, the probability that a signal is of astrophysical origin [cite],  for all the high-mass (systems with total masses in the range of $55-500$ \msun) candidate gravitational-wave signals from compact binary coalescences in the detector data recorded during aLIGO's second observing run (O2).  We find that (a) our search does not identify any new stellar mass or intermediate mass black holes; (b) high-mass events reported in the GWTC-1, including GW170729 (an event with disputed $p_\text{astro}$ amongst various search pipelines) have high \pastro; and that (c) high-mass events detected from independent groups have low \pastro.

\section{Method}

% PyCBC asks for the 5 citations:
% http://pycbc.org/pycbc/latest/html/credit.html#searches-for-compact-binary-coalescence
Following \citet{bcr_paper}, the \bcr is an odds ratio of the coherent signal hypotheses $mathcal{H_S}$ and the incoherent instrumental feature hypothesis $mathcal{H_I}$. It is given by
\begin{equation}
BCR = \frac{\alpha Z^S}{\Pi^D_{i=1}[\beta Z^G_i + (1-\beta)Z^N_i]}\ .
\end{equation}
The numerator is the probability of obtaining a coherent signal, where $Z^S$ is the evidence for $\mathcal{H_S}$ and $\alpha$ is the prior odds for obtaining a signal $\alpha=P(\mathcal{H_S})/P(\mathcal{H_I})$. The denominator is the probability of obtaining an incoherent glitch or noise amongst the network of $D$ detectors, where $Z_G_i$ and $Z_N_i$ are the evidence for obtaining a glitch in detector $i$ and the evidence for obtaining just noise in detector $i$.

We use a Bayesian framework to calculate the \bcr for high-mass candidate events, simulated signal events and  background events (signal-free events ) supplied by the \pycbc \cite{pycbc_code} search in O2  \cite{pycbc_og0, pycbc_og1, pycbc_og2, pycbc_og3, pycbc_og4, pycbc_og5}.  At its core, \pycbc performs a matched-filter search for binary merger signals using a template-bank of gravitational-wave template waveforms to generate triggers. For the trigger to be considered a  candidate event by \pycbc, it must be observed between detectors with the same template and with a time of arrival difference less than the gravitational-wave travel time between detectors \cite{pycbc_og0}. The simulated signal events are artificially created by injecting signals into the detector data. Similarly, \pycbc artificially constructs the background, signal-free, events by applying relative offsets, or time-shifts, between the data of different detectors \cite{pycbc_og0}.

Figure~\ref{fig:templateBank} depicts the \pycbc template bank, the high-mass search space and the candidate events, simulated events and background events from a one-week stretch of \pycbc's search in O2.

Week long stretches of analysis because noise-non stationary, glitches will occur at different rates depending on state of detectors

Dont know the prior odds,

\begin{itemize}
	\item briefly explain PyCBC search
	\item define triggers
	\item Extract PyCBC foreground, background and software injection trigger's GPS times based on template duration of triggers
	\item analyse the triggers using Bayesian Parameter Estimation to determine $\log{Z_s}$, $\log{Z_g}$ and $\log{Z_n}$
	\item reweight results to account for uncertainties in PSDs
	\item Calculate BCR
	\item Calculate PDFs
	\item Find $\alpha$ and $\beta$ values such that separation of background and software injection PDFs is maximally apart.
	\item use tuned $\alpha$ and $\beta$ from previous step to calculate BCRs for foreground
	\item calculate the $p_{astro}$ of the foreground events to determine the probability of the trigger being from an astrophysical source. 
\end{itemize}

@@
